Project data cleaning

### 2024-03-19 R/eda-code/eda.qmd
change dataset 1

---
title: "Project data cleaning"
author: "Xueyan Hu"
date: "2023-02-22"
output: html_document
---

@@

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

@@

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

@@

# Data exploration through figures

Seed type first: bar chart

```{r}
# Calculate average population by seed type
mydata_seed <- mydata %>%
  group_by(Seed) %>%
  summarize(mean_Selective = mean(Selective),
            sd_Selective = sd(Selective))

# Create the bar chart with standard deviation lines
bp1 <- ggplot(mydata_seed, aes(x = Seed, y = mean_Selective, fill = Seed)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7, width = 0.4)  +
  geom_errorbar(data = mydata_seed, aes(x = Seed, ymin = mean_Selective - sd_Selective, ymax = mean_Selective + sd_Selective, color = "black"),
                position = position_dodge(width = 0.1), width = 0.05) +
  labs(title = "The population difference by seed type",
       x = "Seed Type",
       y = "Average Population") +
  scale_color_manual(values = c("Alfalfa" = "lightblue", "Fenugreek" = "pink")) +
  theme_minimal()
plot(bp1)
figure_file = here("results","figures", "seed_type.png")
ggsave(filename = figure_file, plot=bp1) 
```

>>>>

---
title: "Project eda"
author: "Xueyan Hu"
date: "2023-02-22"
output: html_document
---

@@

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(maps)
library(gt)
```

@@

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata_cancer.rds")
#load data
mydata <- readRDS(data_location)
```

# bar chart for crude prevalence according to states

```{r}
# Group by state and calculate the total population and cancer cases
state_total <- mydata %>%
  group_by(StateDesc) %>%
  summarize(TotalPopulation = sum(PopulationCount),
            TotalCancerCases = sum(Cancer_Case))

# Calculate the prevalence for each state
state_total <- state_total %>%
  mutate(TotalPrevalence = TotalCancerCases / TotalPopulation)

# View the resulting dataframe
print(state_total)

# Load the map data for the US states
us_states <- map_data("state")
print(us_states)

# Convert StateDesc column to lowercase
state_total$StateDesc <- tolower(state_total$StateDesc)

# Merge the map data with the state_totals dataframe
map_data <- merge(us_states, state_total, by.x = "region", by.y = "StateDesc", all.x = TRUE)

# Plot the map
map <- ggplot() +
  geom_map(data = map_data, map = map_data,
           aes(x = long, y = lat, map_id = region, fill = TotalPrevalence),
           color = "black", size = 0.1) +
  scale_fill_gradient(name = "TotalPrevalence", low = "lightblue", high = "darkblue") +
  labs(title = "Cancer Prevalence by State") +
  theme_void() +
  theme(legend.position = "right")
plot(map)
```

# List 10 cities that has the highest cancer prevalence

```{r}
top_10_highest <- mydata %>%
  top_n(10, Data_Value) %>%
  select(Year, StateAbbr, CityName, Data_Value)

print(top_10_highest)

# create a table
table <- top_10_highest %>%
  gt() %>%
  tab_spanner(
    label = md('**Year**'),
    columns = c('Year')
  ) %>%
  tab_spanner(
    label = md('**State**'),
    columns = c('StateAbbr')
  ) %>%
  tab_spanner(
    label =  md('**Cancer Prevalence**'),
    columns = c('Data_Value')
  ) %>%
  tab_header(
    title = '2017 top 10 cities which have highest cancer rates'
  ) %>%
  opt_stylize(style = 6, color = 'gray')
```

@@

# fit a simple linear model

```{r}
# create boxplot of Fare by Passenger class
bp1 <- ggplot(mydata, aes(x = as.factor(Pclass), y = Fare)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ticket Fare by Passenger Class",
       x = "Passenger Class",
       y = "Fare") +
  theme_minimal()
plot(bp1)
figure_file = here("results","figures", "seed_type.png")
ggsave(filename = figure_file, plot=bp1) 
```

### 2024-03-19 R/processing-code/processingfile-v1.qmd
change dataset 1

Load needed packages. make sure they are installed.

@@

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","raw_data.xlsx")
rawdata <- readxl::read_excel(data_location)
```

# Check data

First we can look at the codebook

```{r}
codebook <- readxl::read_excel(data_location, sheet ="Codebook")
print(codebook)
```

@@

@@

# Cleaning

The data looks generally clear before cleaning. So I just want to make sure that "Day" could be defined as a categorical variable but not a numeric variable.

```{r}
data1 <- rawdata %>%
  mutate(Day = paste0("Day", substring(Day, 2)))
skimr::skim(data1)
head(data1)
```

# Save data

@@

```{r}
processeddata <- data1
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

>>>>

### load packages

@@

### loading data

```{r}
data_location <- here::here("data","raw-data","500_Cities__Cancer__excluding_skin_cancer__among_adults_aged___18_years_20240314.csv")
rawdata <- read.csv(data_location)
```

@@

The meaning of each variable can be checked [online](https://www.kaggle.com/datasets/yasserh/titanic-dataset/data).

@@

### data cleaning

First, only crude prevalence is interested since we don't know the age structure here. And the prevalence of each city is the summary of all the census tracts from this city, so City will be selected only for further analysis. There are also 2 summed data including all the US prevalence that will not be used for analysis.

```{r}
d1 <- rawdata %>%
  filter(GeographicLevel == "City", DataValueTypeID == "CrdPrv", StateAbbr != "US")
```

Dropping uninterested variables for a smaller dataset.

```{r}
d2 <- d1 %>%
  select(Year, StateAbbr, StateDesc, CityName, DataValueTypeID, Data_Value,PopulationCount)
```

Generating a new variable in case it will be needed later.

```{r}
d3 <- d2 %>%
  mutate(Cancer_Case = round((Data_Value / 100) * PopulationCount))
```

Re-check with the data

```{r}
dplyr::glimpse(d3)
summary(d3)
head(d3)
skimr::skim(d3)
```

### Save data

@@

```{r}
processeddata <- d3
save_data_location <- here::here("data","processed-data","processeddata_cancer.rds")
saveRDS(processeddata, file = save_data_location)
```
